Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
More index items

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

@@

\chapter{CXL Values}
\label{ch:cxlvalues}

@@

\chapter{The CXL Machine}
\label{ch:cxlmachine}

@@

\chapter{Critical Sections}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label}, identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

Mutual exclusion is an example of a \emph{safety property}, a
property that ensures that \emph{nothing bad will happen}, in this case
two processes being in the critical section.
What we need now a \emph{liveness property}: we want to ensure that
\emph{eventually something good will happen}.
There are various possible liveness properties we could use,
but here we will propose the following informally: if some set of processes
$S$ are trying to enter the critical section and any process already in the
intersection eventually leaves, then eventually one process in $S$ will enter
the critical section.
We call this \emph{progress}.

@@

Figure~\ref{fig:mutex} show a high-level state diagram of the code in
Figure~\ref{fig:cs}.
The circles represent states, while the arrows represent possible state
transitions.  The label in each circle summarizes the state;  the label
on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate.

@@

You may already have heard of the concept of a \emph{lock} and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.  Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace} is a
sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable} if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

To solve this problem, we will use what is called an
\emph{inductive invariant}.
An inductive invariant $I$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $I$ holds in the initial state.
\item For any state in which $I$ holds and any process in the
state takes a step, then $I$ also holds in the resulting state.
\end{itemize}
Unlike an invariant, an inductive invariant must also hold over unreachable states.

@@

\chapter{CXL Methods and Pointers}

@@

\chapter{Spinlock}

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{Verbatim}
\end{code}

Many CPUs have an atomic ``test-and-set'' operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{s} points to a shared boolean variable and \texttt{p}
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
It is called \emph{spinlock}, because each process is ``spinning'' until
it can acquire the lock.
The program uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

@@

\begin{quote}
Try doing mutual exclusion with other interlock instructions!
\begin{itemize}
\item Find a description for \texttt{compare-and-swap} on the internet and
update Figure~\ref{fig:tas}, replacing \texttt{tas}.
\item Other interesting interlock instructions include \texttt{swap} and
\texttt{fetch-and-add}.
\end{itemize}
\end{quote}

\chapter{Locks and the Synch Module}

@@

We call a process \emph{blocked} in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

\chapter{Reader/Writer Locks using Busy Waiting}

@@

\chapter{Deadlock}

@@

When multiple processes are synchronizing access to shared resources, they
may end up in a \emph{deadlock} situation where one or more of the processes
end up being blocked indefinitely because each is waiting for another to give
up a resource.
The famous Dutch computer scientist Edsger W.~Dijkstra illustrated this using
a scenario he called ``Dining Philosophers.''

@@

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P} operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V} operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.

@@

\chapter{Bounded Buffer}

@@

\chapter{Split Binary Semaphores}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections} (CCSs)~\cite{Hoare73}.
All critical sections are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

\chapter{Starvation}
\label{ch:starvation}

@@

\chapter{Non-Blocking Synchronization}

@@

\chapter{Barrier Synchronization}

@@

\chapter{Networking: Alternating Bit Protocol}

@@

The \texttt{bag} module has various useful methods that operate on bags or
multisets:

@@

The \texttt{list} module has various useful methods that operate on lists
or tuples:

@@

A \emph{micro step} is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step}
\glsadd{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.

>>>>

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

What is going on is that the CXL program is compiled to machine instructions,
\index{machine instruction}
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

@@

\chapter{CXL Values}
\label{ch:cxlvalues}
\index{CXL Value}

@@

\chapter{The CXL Machine}
\label{ch:cxlmachine}
\index{CXL machine}

@@

\chapter{Critical Sections}
\index{critical section}
\index{critical region}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label}, identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

Mutual exclusion is an example of a \emph{safety property},
\index{safety property}
a property that ensures that \emph{nothing bad will happen}, in this case
two processes being in the critical section.
What we need now a \emph{liveness property}:
\index{liveness property}
we want to ensure that
\emph{eventually something good will happen}.
There are various possible liveness properties we could use,
but here we will propose the following informally: if some set of processes
$S$ are trying to enter the critical section and any process already in the
intersection eventually leaves, then eventually one process in $S$ will enter
the critical section.
We call this \emph{progress}.
\index{progress}

@@

Figure~\ref{fig:mutex} show a high-level state diagram
\index{state diagram}
of the code in
Figure~\ref{fig:cs}.
The circles represent states, while the arrows represent possible state
transitions (high-level steps).  The label in each circle summarizes the state;
the label on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}
and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace} is a
sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable} if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

To solve this problem, we will use what is called an
\emph{inductive invariant}.
\index{inductive invariant}
An inductive invariant $I$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $I$ holds in the initial state.
\item For any state in which $I$ holds and any process in the
state takes a step, then $I$ also holds in the resulting state.
\end{itemize}
Unlike an invariant, an inductive invariant must also hold over unreachable states.

@@

\chapter{CXL Methods and Pointers}
\index{CXL method}

@@

\chapter{Spinlock}
\index{spinlock}

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{Verbatim}
\end{code}

Many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}
\index{TAS}
operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{s} points to a shared boolean variable and \texttt{p}
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
It is called \emph{spinlock},
\index{spinlock}
because each process is ``spinning'' until
it can acquire the lock.
The program uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

@@

\begin{quote}
Try doing mutual exclusion with other interlock instructions!
\begin{itemize}
\item Find a description for \texttt{compare-and-swap} (CAS)
\index{compare-and-swap}
\index{CAS}
on the internet and
update Figure~\ref{fig:tas}, replacing \texttt{tas}.
\item Other interesting interlock instructions include \texttt{swap} and
\texttt{fetch-and-add}.
\index{fetch-and-add}
\end{itemize}
\end{quote}

\chapter{Locks and the Synch Module}
\index{lock}
\index{synch module}

@@

We call a process \emph{blocked}
\index{blocked}
in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

\chapter{Reader/Writer Locks using Busy Waiting}
\index{reader/writer lock}
\index{busy waiting}

@@

\chapter{Deadlock}
\index{deadlock}

@@

When multiple processes are synchronizing access to shared resources, they
may end up in a \emph{deadlock} situation where one or more of the processes
end up being blocked indefinitely because each is waiting for another to give
up a resource.
The famous Dutch computer scientist Edsger W.~Dijkstra illustrated this using
a scenario he called ``Dining Philosophers.''
\index{dining philosopher}

@@

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P}
\index{P}
\index{Procure}
operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V}
\index{V}
\index{Vacate}
operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.

@@

\chapter{Bounded Buffer}
\index{bounded buffer}
\index{producer/consumer problem}

@@

\chapter{Split Binary Semaphores}
\index{split binary semaphore}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections}
\index{conditional critical section}
(CCSs)~\cite{Hoare73}.
All critical sections are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

\chapter{Starvation}
\label{ch:starvation}
\index{starvation}

@@

\chapter{Non-Blocking Synchronization}
\index{non-blocking synchronization}

@@

\chapter{Barrier Synchronization}
\index{barrier synchronization}

@@

\chapter{Networking: Alternating Bit Protocol}
\index{alternating bit protocol}

@@

The \texttt{bag} module
\index{bag module}
has various useful methods that operate on bags or
multisets:

@@

\noindent
\index{list module}
The \texttt{list} module has various useful methods that operate on lists
or tuples:

@@

A \emph{micro step}
\index{micro step}
is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step}
\index{macro step}
\glsadd{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.
